plyr: Split-Apply-Combine for Mortals

### 2013-12-02 _knitr-posts/2013-12-01-plyr.Rmd
Add revised plyr intro post

@@

“`plyr` is a set of tools that solves a common set of problems: you need
to break a big problem down into manageable pieces, operate on each
piece and then put all the pieces back together. It’s already possible
to do this with split and the apply functions, but `plyr` just makes it
all a bit easier…”

This is a very quick introduction to `plyr`. For more details see Hadley
Wickham’s introductory guide [The split-apply-combine strategy for data
analysis](http://www.jstatsoft.org/v40/i01) (2011, Journal of
Statistical Software, Vol 40). There’s quite a bit of discussion online
in general, and especially on
[stackoverflow.com](http://stackoverflow.com/questions/tagged/plyr).

@@

1.  The code is cleaner (once you’re familiar with the concept). The
    code can be easier to code and read, and less error prone because you
    don’t have to deal with subsetting and you don’t have to deal with saving
    your results.

@@

The basic format is 2 letters followed by `ply()`. The first letter
refers to the format in and the second to the format out.

The 3 main letters are:

@@

I’ve ignored some less common format options:

@@

`plyr` provides a consistent and easy-to-work-with format for `apply`
functions with control over the input and output formats. Some of the
functionality can be duplicated with base R functions (but with less
consistent syntax). Also, few R `apply` functions work directly with
data frames as input and output and data frames are a common object
class to work with.

Base R `apply` functions (from a presentation given by Hadley):

@@

Let’s take a simple example. We’ll take a data frame, split it up by
`year`, calculate the coefficient of variation of the `count`, and
return a data frame. This could easily be done on one line, but I’m
expanding it here to show the format a more complex function could take.

@@

It is often convenient to use these functions within `plyr`. `transform`
acts as it would normally as the base R function and modifies an
existing data frame. `summarise` creates a new condensed data
frame.

@@

Bonus function: `mutate`. `mutate` works like `transform` but lets you
build on columns you build.

@@

You can use `plyr` to plot data by throwing away the output with an
underscore (`_`). This is a bit cleaner than a for loop since you don’t
have to subset the data manually.

@@

`plyr` can be slow — particularly if you are working with very large
datasets that involve a lot of subsetting. Hadley is working on this and
an in-development version of `plyr`, `dplyr`, can run much faster (<https://github.com/hadley/dplyr>).
However, it’s important to remember that typically the speed that you
can write code and understand it later is the rate-limiting step.

@@

Use a base R `apply` function:

>>>>

*Along with my [earlier post](http://localhost:4000/2013/10/19/reshape.html) on the `reshape2` package, I will continue to post my course notes from Data Wrangling and Visualization in R, a graduate-level course I co-taught last semester at Simon Fraser University.*

@@

"`plyr` is a set of tools that solves a common set of problems: you need
to break a big problem down into manageable pieces, operate on each
piece and then put all the pieces back together. It's already possible
to do this with split and the apply functions, but `plyr` just makes it
all a bit easier..."

This is a very quick introduction to `plyr`. For more details see Hadley
Wickham's introductory guide [The split-apply-combine strategy for data
analysis](http://www.jstatsoft.org/v40/i01). There's quite a bit of discussion
online in general, and especially on
[stackoverflow.com](http://stackoverflow.com/questions/tagged/plyr).

@@

1.  The code is cleaner (once you're familiar with the concept). The
    code can be easier to code and read, and less error prone because you
    don't have to deal with subsetting and you don't have to deal with saving
    your results.

@@

The basic format is two letters followed by `ply()`. The first letter
refers to the format in and the second to the format out.

The three main letters are:

@@

I've ignored some less common format options:

@@

`plyr` provides a consistent and easy-to-work-with format for `apply`
functions with control over the input and output formats. Some of the
functionality can be duplicated with base `R` functions (but with less
consistent syntax). Also, few `R` `apply` functions work directly with
data frames as input and output and data frames are a common object
class to work with.

Base `R` `apply` functions (from a presentation given by Hadley Wickham):

@@

Let's take a simple example. We'll take a data frame, split it up by
`year`, calculate the coefficient of variation of the `count`, and
return a data frame. This could easily be done on one line, but I'm
expanding it here to show the format a more complex function could take.

@@

It is often convenient to use these functions within one of the `**ply`
functions. `transform` acts as it would normally as the base `R` function and
modifies an existing data frame. `summarise` creates a new condensed data
frame.

@@

Bonus function: `mutate`. `mutate` works like `transform` but lets you
build on columns.

@@

You can use `plyr` to plot data by throwing away the output with an
underscore (`_`). This is a bit cleaner than a for loop since you don't
have to subset the data manually.

@@

`plyr` can be slow --- particularly if you are working with very large
datasets that involve a lot of subsetting. Hadley is working on this and
an in-development version of `plyr`, `dplyr`, can run much faster (<https://github.com/hadley/dplyr>).
However, it's important to remember that typically the speed that you
can write code and understand it later is the rate-limiting step.

@@

Use a base `R` `apply` function:
